---
title: "Simulate datasets with batch effect"
author: 
  - name: Dongyuan Song
    affiliation:
    - Bioinformatics IDP, University of California, Los Angeles
    email: dongyuansong@ucla.edu
  - name: Qingyang Wang
    affiliation:
    - Department of Statistics, University of California, Los Angeles
    email: qw802@g.ucla.edu
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('scDesign3')`"
vignette: >
  %\VignetteIndexEntry{scDesign3-batchEffect-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{css, echo=FALSE}
pre {
  white-space: pre !important;
  overflow-x: scroll !important;
}
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE}
library(scDesign3)
library(ggplot2)
```


## Introduction
In this tutorial, we will show how to use scDesign3 to simulate data with original batch effects and how to remove the batch effects. We will also demostrate how to add ariticial batch effects.

## Read in the reference data
The raw data is from the `SeuratData` package. The data is called `pbmcsca` in the package; it is PBMC Systematic Comparative Analysis dataset from the Broad Institute.

```{r, message=FALSE}
example_sce <- readRDS((url("https://www.dropbox.com/sh/0tptrft7hsstpi3/AACFsTJvmzgBKicMsbFzxbyUa/BATCH_sce.rds?raw=1")))
print(example_sce)
```
To save computational time, we only use the top 100 genes.
```{r, message=FALSE}
example_sce <- example_sce[1:100, ]
```

The column `batch` in this example dataset's colData contains the batch information.
```{r, message=FALSE}
head(colData(example_sce))
```


## Simulation
We can simulate a new data with batch effect information.
```{r, message=FALSE, warning=FALSE, results='hide'}
simu_res <- scdesign3(sce = example_sce, 
                              assay_use = "counts", 
                              celltype = "cell_type", 
                              pseudotime = NULL, 
                              spatial = NULL, 
                              other_covariates = c("batch"), 
                              mu_formula = "cell_type + batch", 
                              sigma_formula = "1", 
                              family_use = "nb", 
                              n_cores = 2, 
                              usebam = FALSE, 
                              corr_formula = "1", 
                              copula = "gaussian", 
                              DT = TRUE, 
                              pseudo_obs = FALSE, 
                              return_model = FALSE)
```


We can also remove the batch effect and generate new data.
```{r, message=FALSE, warning=FALSE, results='hide'}
BATCH_data <- construct_data(
  sce = example_sce,
  assay_use = "counts",
  celltype = "cell_type",
  pseudotime = NULL,
  spatial = NULL,
  other_covariates = c("batch"),
  corr_by = "1"
)
```
```{r, message=FALSE, warning=FALSE, results='hide'}
BATCH_marginal <- fit_marginal(
  data = BATCH_data,
  predictor = "gene",
  mu_formula = "cell_type + batch",
  sigma_formula = "1",
  family_use = "nb",
  n_cores = 2,
  usebam = FALSE
)
```
```{r, message=FALSE, warning=FALSE, results='hide'}
BATCH_copula <- fit_copula(
    sce = example_sce,
    assay_use = "counts",
    marginal_list = BATCH_marginal,
    family_use = "nb",
    copula = "gaussian",
    n_cores = 2,
    new_covariate = NULL,
    input_data = BATCH_data$dat
  )
```
```{r, message=FALSE, warning=FALSE, results='hide'}
BATCH_marginal_null <- lapply(BATCH_marginal, function(x) {
  x$coefficients[length(x$coefficients)] <- 0
  x
})
```
```{r, message=FALSE, warning=FALSE, results='hide'}
BATCH_para_null <- extract_para(
    sce = example_sce,
    marginal_list = BATCH_marginal_null,
    n_cores = 2,
    family_use = "nb",
    new_covariate = NULL
  )
```
```{r, message=FALSE, warning=FALSE, results='hide'}
BATCH_newcount_null <- simu_new(
    sce = example_sce,
    mean_mat = BATCH_para_null$mean_mat,
    sigma_mat = BATCH_para_null$sigma_mat,
    zero_mat = BATCH_para_null$zero_mat,
    quantile_mat = NULL,
    copula_list = BATCH_copula$copula_list,
    n_cores = 2,
    family_use = "nb",
    input_data = BATCH_data$dat,
    new_covariate = BATCH_data$new_covariate
  )
```

Additionally, we can alter the batch effect information and simulate new dataset with altered batch effect information.
```{r, message=FALSE, warning=FALSE, results='hide'}
BATCH_marginal_alter <- lapply(BATCH_marginal, function(x) {
  x$coefficients[length(x$coefficients)] <- rnorm(1, mean = 5, sd = 2)
  x
})
```
```{r, message=FALSE, warning=FALSE, results='hide'}
BATCH_para_alter <- extract_para(
    sce = example_sce,
    marginal_list = BATCH_marginal_alter,
    n_cores = 2,
    family_use = "nb",
    new_covariate = NULL
  )
```
```{r, message=FALSE, warning=FALSE, results='hide'}
BATCH_newcount_alter <- simu_new(
    sce = example_sce,
    mean_mat = BATCH_para_alter$mean_mat,
    sigma_mat = BATCH_para_alter$sigma_mat,
    zero_mat = BATCH_para_alter$zero_mat,
    quantile_mat = NULL,
    copula_list = BATCH_copula$copula_list,
    n_cores = 2,
    family_use = "nb",
    input_data = BATCH_data$dat,
    new_covariate = BATCH_data$new_covariate
  )
```

## Visulization
We first create three `SingleCellExperiment` objects using three count matrices generated above.
```{r, message=FALSE}
simu_res_list <- lapply(list(simu_res$new_count, BATCH_newcount_null, BATCH_newcount_alter), function(x){
  simu_sce <- example_sce
  counts(simu_sce) <- x
  logcounts(simu_sce) <- log1p(counts(simu_sce))
  return(simu_sce)
})
```
Then, we visualize the results.
```{r, message=FALSE, fig.width=12}
set.seed(123)
compare_figure <- plot_reduceddim(ref_sce = example_sce, 
                                  sce_list = simu_res_list, 
                                  name_vec = c("Reference", "w/ Batch", "w/o Batch","Aritifical Batch"),
                                  assay_use = "logcounts", 
                                  if_plot = TRUE, 
                                  color_by = "cell_type", 
                                  shape_by = "batch",
                                  n_pc = 20)
plot(compare_figure$p_umap)
```
