---
title: "Simulate datasets with condition effect"
author: 
  - name: Dongyuan Song
    affiliation:
    - Bioinformatics IDP, University of California, Los Angeles
    email: dongyuansong@ucla.edu
  - name: Qingyang Wang
    affiliation:
    - Department of Statistics, University of California, Los Angeles
    email: qw802@g.ucla.edu
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('scDesign3')`"
vignette: >
  %\VignetteIndexEntry{scDesign-conditionEffect-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{css, echo=FALSE}
pre {
  white-space: pre !important;
  overflow-x: scroll !important;
}
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE}
library(scDesign3)
```

## Introduction
In this tutorial, we will show how to use scDesign3 to simulate data with condition effects and how to adjust the condition effects.

## Read in the reference data
The raw data is from the `SeuratData` package. The data is called `ifnb` in the package; it is PBMC data simulated and controlled by IFNB.

```{r, message=FALSE}
example_sce <- readRDS((url("https://www.dropbox.com/sh/0tptrft7hsstpi3/AABoozvjRadAygeSSJFoc2OOa/IFNB_sce.rds?raw=1")))
print(example_sce)
```
To save computational time, we only use the top 100 genes.
```{r, message=FALSE}
example_sce <- example_sce[1:100,]
```


The condition information is stored in colData of the example dataset.
```{r, message=FALSE}
head(colData(example_sce))
```
## Simulation
First, we will simulate new data with the condition effects.
```{r, message=FALSE}
simu_res <- scdesign3(
    sce = example_sce,
    assay_use = "counts",
    celltype = "cell_type",
    pseudotime = NULL,
    spatial = NULL,
    other_covariates = c("condition"),
    mu_formula = "cell_type + condition + cell_type*condition",
    sigma_formula = "cell_type",
    family_use = "nb",
    n_cores = 2,
    usebam = FALSE,
    corr_formula = "cell_type",
    copula = "gaussian",
    DT = TRUE,
    pseudo_obs = FALSE,
    return_model = FALSE,
    nonzerovar = FALSE
  )


```

Then, we can also simulate a new dataset with condition effects on B cells removed.
```{r, message=FALSE}
IFNB_data <- construct_data(
  sce = example_sce,
  assay_use = "counts",
  celltype = "cell_type",
  pseudotime = NULL,
  spatial = NULL,
  other_covariates = c("condition"),
  corr_by = "cell_type"
)
```
```{r, message=FALSE}
IFNB_marginal <- fit_marginal(
  data = IFNB_data,
  predictor = "gene",
  mu_formula = "cell_type + condition + cell_type*condition",
  sigma_formula = "cell_type",
  family_use = "nb",
  n_cores = 2,
  usebam = FALSE
)
```
```{r, message=FALSE}
IFNB_copula <- fit_copula(
    sce = example_sce,
    assay_use = "counts",
    marginal_list = IFNB_marginal,
    family_use = "nb",
    copula = "gaussian",
    n_cores = 2,
    new_covariate = NULL,
    input_data = IFNB_data$dat
  )
```
```{r, message=FALSE}
IFNB_marginal_null_B <- lapply(IFNB_marginal, function(x) {
  x$coefficients["cell_typeB:conditionSTIM"] <- x$coefficients["conditionSTIM"]
  x
})
```
```{r, message=FALSE}
IFNB_para_null_B <- extract_para(
    sce = example_sce,
    marginal_list = IFNB_marginal_null_B,
    n_cores = 2,
    family_use = "nb",
    new_covariate = NULL
  )
```
```{r, message=FALSE}
IFNB_newcount_null_B <- simu_new(
    sce = example_sce,
    mean_mat = IFNB_para_null_B$mean_mat,
    sigma_mat = IFNB_para_null_B$sigma_mat,
    zero_mat = IFNB_para_null_B$zero_mat,
    quantile_mat = NULL,
    copula_list = IFNB_copula$copula_list,
    n_cores = 2,
    family_use = "nb",
    input_data = IFNB_data$dat,
    new_covariate = IFNB_data$new_covariate
  )
```

## Visulization
We first create two `SingleCellExperiment` objects using two count matrices generated above.
```{r, message=FALSE}
simu_res_list <- lapply(list(simu_res$new_count,IFNB_newcount_null_B ), function(x){
  logcounts(example_sce) <- log1p(counts(example_sce))
  simu_sce <- example_sce
  counts(simu_sce) <- x
  logcounts(simu_sce) <- log1p(counts(simu_sce))
  return(simu_sce)
})
logcounts(example_sce) <- log1p(counts(example_sce))
```
Then, we visualize the results.
```{r, message=FALSE, fig.width=12}
set.seed(123)
compare_figure <- plot_reduceddim(ref_sce = example_sce, 
                                  sce_list = simu_res_list, 
                                  name_vec = c("Reference", "w/ Condition", "Modified Condition"),
                                  assay_use = "logcounts", 
                                  if_plot = TRUE, 
                                  color_by = "cell_type", 
                                  shape_by = "condition",
                                  n_pc = 20)
plot(compare_figure$p_umap)
```



