---
title: "Simulate datasets with cell type and modified cell-type proportions "
author: 
  - name: Dongyuan Song
    affiliation:
    - Bioinformatics IDP, University of California, Los Angeles
    email: dongyuansong@ucla.edu
  - name: Qingyang Wang
    affiliation:
    - Department of Statistics, University of California, Los Angeles
    email: qw802@g.ucla.edu
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('scDesign3')`"
vignette: >
  %\VignetteIndexEntry{scDesign3-cellType-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---
```{css, echo=FALSE}
pre {
  white-space: pre !important;
  overflow-x: scroll !important;
}
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    message = FALSE,
    collapse = TRUE,
    comment = "#>",
    crop = NULL ## Related to https://stat.ethz.ch/pipermail/bioc-devel/2020-April/016656.html
)
tools::R_user_dir("scDesign3Workflow", which="cache")
```

```{r, message=FALSE, warning=FALSE, results='hide'}
library(scDesign3)
library(SingleCellExperiment)
library(ggplot2)
library(DuoClustering2018)
library(scran)
library(tidyverse)
theme_set(theme_bw())
```

## Introduction
In this tutorial, we will demonstrate how to use scDesign3 to simulate datasets with cell-type labels as covariates. We will also guide you through modifying cell-type proportions and generating corresponding synthetic data using our step-by-step functions.

## Read in the reference data
The raw data is from the R package `DuoClustering2018` which contain a set of datasets with true cell type labels. 
```{r, message=FALSE}
sce <- get("sce_filteredExpr10_Zhengmix4eq")(metadata = FALSE)
colData(sce)$cell_type = as.factor(colData(sce)$phenoid)
```

To save computational time, we only use the top 200 highly variable genes.
```{r}
ngene <- 200
logcounts(sce) <- log1p(counts(sce))
temp_sce <- modelGeneVar(sce)
chosen <- getTopHVGs(temp_sce, n = ngene)
sce <- sce[chosen,]
```

## Simulation with original cell-type labels
Then, we set the parameter `celltype = "cell_type"`, indicating that the cell-type labels are stored in the `cell_type` column within the `colData` of the input `SingleCellExperiment` object. We also set the parameters `mu_formula`  and `sigma_formula` as `cell_type`, specifying that in the GAMLSS model, both the log-transformed mean and dispersion parameters for each gene’s marginal distribution (in this case, the negative binomial distribution) have a linear relationship with the covariate `cell_type`. The parameter `corr_formula` is set to `cell_type`,  meaning that each cell type will have its own unique gene-gene correlation structure.
```{r, message=FALSE, warning=FALSE, results='hide'}
set.seed(123)
example_simu <- scdesign3(
    sce = sce,
    assay_use = "counts",
    celltype = "cell_type",
    pseudotime = NULL,
    spatial = NULL,
    other_covariates = NULL,
    mu_formula = "cell_type",
    sigma_formula = "cell_type",
    family_use = "nb",
    n_cores = 2,
    usebam = FALSE,
    corr_formula = "cell_type",
    copula = "gaussian",
    DT = TRUE,
    pseudo_obs = FALSE,
    return_model = FALSE,
    nonzerovar = FALSE,
    parallelization = "pbmcmapply"
  )
```

Then, we can create the `SinglecellExperiment` object using the synthetic count matrix and store the `logcounts` to the input and synthetic `SinglecellExperiment` objects.
```{r}
logcounts(sce) <- log1p(counts(sce))
simu_sce <- SingleCellExperiment(list(counts = example_simu$new_count), colData = example_simu$new_covariate)
logcounts(simu_sce) <- log1p(counts(simu_sce))
```


## Simulation with modified cell-type proportions
To change the cell-type proportions in the simulated dataset, we need to use scDesign3’s step-by-step functions. Specifically, we provide a new set of cell-type labels with the desired proportions in the `extract_para()` and `simu_new()` functions. The detailed code is shown below.

### 1. Construct the input dataset.
```{r, message=FALSE, warning=FALSE, results='hide', eval=TRUE}
example_data <- construct_data(
    sce = sce,
    assay_use = "counts",
    celltype = "cell_type",
    pseudotime = NULL,
    spatial = NULL,
    other_covariates = NULL,
    corr_by = "cell_type"
  )
```
### 2. Fit regression models for each feature based on your specification. 
Here, we set the parameters `mu_formula`  and `sigma_formula` as `cell_type`, indicating that in the GAMLSS model, the log-transformed mean and dispersion parameters for each gene's marginal distribution (in this case, the negative binomial distribution) has a linear relationship with the covariate `cell_type`.
```{r, message=FALSE, warning=FALSE, results='hide', eval=TRUE}
example_marginal <- fit_marginal(
    data = example_data,
    predictor = "gene",
    mu_formula = "cell_type",
    sigma_formula = "cell_type",
    family_use = "nb",
    n_cores = 2,
    usebam = FALSE,
    parallelization = "pbmcmapply"
  )
```
### 3. Fit a copula, obtain AIC and BIC.
```{r, message=FALSE, warning=FALSE, results='hide', eval=TRUE}
set.seed(123)
example_copula <- fit_copula(
    sce = sce,
    assay_use = "counts",
    marginal_list = example_marginal,
    family_use = "nb",
    copula = "gaussian",
    n_cores = 2,
    input_data = example_data$dat
  )
```
### 4. Extract out the estimated parameters so you can make some modifications and use the modified parameters to generate new data if needed.

The cell-type proportions in the inputted dataset are shown below
```{r}
table(colData(sce)$cell_type) / dim(sce)[2]
```

For demonstration purpose, we generate a set of cell-type labels with 0\% of b cells, 0\% of CD 14 monocytes, 20\% of naive cytotoxic cell, and 80\% regulartory T cells. We will supply the data frame `new_ct` generated below to the `new_covariate` parameter in `extract_para()` and `simu_new()`. The `new_covariate` parameters must include all the covariates specified in `mu_formula` and `sigma_formula`, and a column called `corr_group` indicating the gene-gene correlation structure. Here, since we specify `corr_formula = "cell_type"` in `construct_data()`, the `corr_group` column in `new_ct` also need to reflect this. 
```{r}
ct_prop <- c(0, 0, 0.2,0.8)
unique_combined <- example_data$dat %>%expand(nesting(cell_type, corr_group))
new_ct <- as.data.frame(lapply(unique_combined, rep,round(ct_prop*dim(sce)[2])))
head(new_ct)
table(new_ct$cell_type)/dim(new_ct)[1]
```
```{r, message=FALSE, warning=FALSE, results='hide', eval=TRUE}
example_para <- extract_para(
    sce = sce,
    marginal_list = example_marginal,
    n_cores = 1,
    family_use = "nb",
    new_covariate = new_ct,
    data = example_data$dat
  )
```

### 5. Simulate the new count matrix.
```{r, message=FALSE, warning=FALSE, results='hide', eval=TRUE}
set.seed(123)
example_newcount <- simu_new(
    sce = sce,
    mean_mat = example_para$mean_mat,
    sigma_mat = example_para$sigma_mat,
    zero_mat = example_para$zero_mat,
    quantile_mat = NULL,
    copula_list = example_copula$copula_list,
    n_cores = 1,
    family_use = "nb",
    input_data = example_data$dat,
    new_covariate = new_ct,
    important_feature = example_copula$important_feature,
    filtered_gene = example_data$filtered_gene
  )
```

Then, we can create the `SinglecellExperiment` object using the synthetic count matrix and store the `logcounts` to the input and synthetic `SinglecellExperiment` objects.
```{r}
logcounts(sce) <- log1p(counts(sce))
simu_sce2 <- SingleCellExperiment(list(counts = example_newcount), colData = data.frame(cell_type = new_ct$cell_type))
logcounts(simu_sce2) <- log1p(counts(simu_sce2))
```

## Visualization
```{r}
set.seed(123)
compare_figure <- plot_reduceddim(ref_sce = sce, 
                                  sce_list = list(simu_sce, simu_sce2), 
                                  name_vec = c("Reference", "Same cell-type proportions", "Modified cell-type proportions"),
                                  assay_use = "logcounts", 
                                  if_plot = TRUE, 
                                  color_by = "cell_type", 
                                  n_pc = 20)
plot(compare_figure$p_umap)
```



## Session information
```{r}
sessionInfo()
```
