---
title: "Parallelization in scDesign3"
author: 
  - name: Dongyuan Song
    affiliation:
    - Bioinformatics IDP, University of California, Los Angeles
    email: dongyuansong@ucla.edu
  - name: Qingyang Wang
    affiliation:
    - Department of Statistics, University of California, Los Angeles
    email: qw802@g.ucla.edu
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('scDesign3')`"
vignette: >
  %\VignetteIndexEntry{scDesign3-parallelization-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{css, echo=FALSE}
pre {
  white-space: pre !important;
  overflow-x: scroll !important;
}
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
tools::R_user_dir("scDesign3", which="cache")
```

```{r setup, message=FALSE}
library(scDesign3)
library(BiocParallel)
```



## Introduction
In this tutorial, we will demonstrate how to use the parallelization functions from three packages: `parallel`, `pbmcapply`,`BiocParallel`. For the demonstration purpose, we will use our one-shot function `scdesign3` to show how to adjust the parameter to use different parallelization functions. Although we only show how to change parallelization functions for our one-shot function `scdesign3`, the same logic holds for all other methods in our package with the parameters `parallelization` and `BPPARAM`. 

* For Linux/Mac users:
The possible parallel methods include `mcmapply`, `pbmcmapply` and `bpmapply`, which are parallelization functions from `parallel`, `pbmcapply`,`BiocParallel`, respectively. If you are using the `bpmapply` method, then you should give the parameter `BPPARAM` in `scdesign3` 
the output from `MulticoreParam` or `SnowParam`, which are two functions from `BiocParallel`.

* For Windows users:
The only choice is to choose `bpmapply` method and run this function by using the `BPPARAM` parameter in `scdesign3` to the output from the `SnowParam` function in `BiocParallel`. **Setting more than 1 core is not allowed in other methods**.


## Read in the reference data
The raw data is from the [scvelo](https://scvelo.readthedocs.io/scvelo.datasets.pancreas/), which describes pancreatic endocrinogenesis. We pre-select the top 1000 highly variable genes and filter out some cell types to ensure a **single trajectory**.
```{r}
example_sce <- readRDS((url("https://figshare.com/ndownloader/files/40581992")))
print(example_sce)
```

To save computational time, we only use the top 100 genes.
```{r}
example_sce <- example_sce[1:100, ]
```


## Paralleization function from `parallel`
The parameter `parallelization` in `scdesign3` is set to use `mcmapply` from the R package `parallel` by default. Therefore, if you want to use the 
parallelization function from `parallel`, you do not need to modify the parameter `parallelization` in `scdesign3`.
```{r, message=FALSE, warning=FALSE}
set.seed(123)
example_simu <- scdesign3(
    sce = example_sce,
    assay_use = "counts",
    celltype = "cell_type",
    pseudotime = "pseudotime",
    spatial = NULL,
    other_covariates = NULL,
    mu_formula = "s(pseudotime, k = 10, bs = 'cr')",
    sigma_formula = "1",
    family_use = "nb",
    n_cores = 2,
    usebam = FALSE,
    corr_formula = "1",
    copula = "gaussian",
    DT = TRUE,
    pseudo_obs = FALSE,
    return_model = FALSE,
    nonzerovar = FALSE
  )
```


## Paralleization function from `pbmcapply`
The `pbmcmapply` function from the R package `pbmcapply` gives users a progress bar of the paralleilization. To use `pbmcmapply`, you need to set the parameter `paralleilization` to "pbmcmapply".
```{r, message=FALSE, warning=FALSE}
set.seed(123)
example_simu <- scdesign3(
    sce = example_sce,
    assay_use = "counts",
    celltype = "cell_type",
    pseudotime = "pseudotime",
    spatial = NULL,
    other_covariates = NULL,
    mu_formula = "s(pseudotime, k = 10, bs = 'cr')",
    sigma_formula = "1",
    family_use = "nb",
    n_cores = 2,
    usebam = FALSE,
    corr_formula = "1",
    copula = "gaussian",
    DT = TRUE,
    pseudo_obs = FALSE,
    return_model = FALSE,
    nonzerovar = FALSE,
    parallelization = "pbmcmapply"
  )
```



## Paralleization function from `BiocParallel`
The `pbmapply` function from the R package `BicoParallel` is another choice of parallelization. To use `pbmapply`, you need to set the parameter `parallelization` to "pbmapply" and set the parameter `BPPARAM` to the output of the function `MulticoreParam()` or `SnowParam()` from `BiocParallel`. More detail documentation about how to set the parameter `BPPARAM` can be found [here](https://www.bioconductor.org/packages/devel/bioc/manuals/BiocParallel/man/BiocParallel.pdf).
```{r, message=FALSE, warning=FALSE}
set.seed(123)
example_simu <- scdesign3(
    sce = example_sce,
    assay_use = "counts",
    celltype = "cell_type",
    pseudotime = "pseudotime",
    spatial = NULL,
    other_covariates = NULL,
    mu_formula = "s(pseudotime, k = 10, bs = 'cr')",
    sigma_formula = "1",
    family_use = "nb",
    n_cores = 2,
    usebam = FALSE,
    corr_formula = "1",
    copula = "gaussian",
    DT = TRUE,
    pseudo_obs = FALSE,
    return_model = FALSE,
    nonzerovar = FALSE,
    parallelization = "pbmapply",
    BPPARAM = MulticoreParam()
  )
```

```{r, message=FALSE, warning=FALSE}
set.seed(123)
example_simu <- scdesign3(
    sce = example_sce,
    assay_use = "counts",
    celltype = "cell_type",
    pseudotime = "pseudotime",
    spatial = NULL,
    other_covariates = NULL,
    mu_formula = "s(pseudotime, k = 10, bs = 'cr')",
    sigma_formula = "1",
    family_use = "nb",
    n_cores = 2,
    usebam = FALSE,
    corr_formula = "1",
    copula = "gaussian",
    DT = TRUE,
    pseudo_obs = FALSE,
    return_model = FALSE,
    nonzerovar = FALSE,
    parallelization = "pbmapply",
    BPPARAM = SnowParam()
  )
```