---
title: "Evaluate clustering goodness-of-fit by scDesign3"
author: 
  - name: Dongyuan Song
    affiliation:
    - Bioinformatics IDP, University of California, Los Angeles
    email: dongyuansong@ucla.edu
  - name: Qingyang Wang
    affiliation:
    - Department of Statistics, University of California, Los Angeles
    email: qw802@g.ucla.edu
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('scDesign3')`"
vignette: >
  %\VignetteIndexEntry{scDesign3-clusterQuality-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{css, echo=FALSE}
pre {
  white-space: pre !important;
  overflow-x: scroll !important;
}
```


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
tools::R_user_dir("scDesign3", which="cache")
```


```{r, message=FALSE}
library(scDesign3)
library(DuoClustering2018)
library(dplyr)
library(scran)
library(aricode)
library(ggplot2)
theme_set(theme_bw())
```

## Introduction
In this tutorial, we will show how to use scDesign3 to evaluate the clustering goodness-of-fit for different cell-type assignments. If the true labels are unavailable and we have little prior knowledge, the scDesign3 BIC can serve as an unsupervised metric.


## Read in the reference data
The raw data is from the R package `DuoClustering2018` which contain a set of datasets with various clustering results.
```{r, message=FALSE}
Zhengmix4eq_sce <- get("sce_filteredExpr10_Zhengmix4eq")(metadata = FALSE)
res <- get("clustering_summary_filteredExpr10_Zhengmix4eq_v1")(metadata = FALSE)
  
res_sub <- res %>% dplyr::filter(method %in% c("SC3", "Seurat", "PCAHC", "PCAKmeans", "CIDR") & run == 1)
res_sub_list <- res_sub %>% group_by(dataset, method, run, k, resolution) %>% group_split()
```

For demonstration purpose, we use the `Zhengmix4eq` dataset in the package with top 100 highly variable genes and the corresponding k-means clustering results with k = $2 ,\cdots, 10$.
```{r, message=FALSE}
kmeans_res <- Filter(function(x){all(x$method == "PCAKmeans")}, res_sub_list)
ncell <- ncol(Zhengmix4eq_sce)
ngene <- 100
zheng_sce <- modelGeneVar(Zhengmix4eq_sce)
chosen <- getTopHVGs(zheng_sce, n = ngene)
sce <- Zhengmix4eq_sce[chosen, ]
ntrain <- round(ncell/2)
  
set.seed(123)
train_index <- sample(seq_len(ncell), ntrain, replace = FALSE)
train_sce <- sce[, train_index]
```

## Simulation
We then use the nine different cell-type clustering information (the results from k-means clustering with k = $2, \cdots, 10$) to simulate nine synthetic datasets. In the code below, we iteratively run scDesign3, and in each iteration, we use a different set of k-means clustering labels as the `celltype` covariate. 
```{r, message=FALSE, warning=FALSE, results='hide', eval=TRUE}
set.seed(123)
scDesign3_result <- lapply(kmeans_res, function(x) {
      dat <- x %>% dplyr::select(c("cell", "cluster")) %>% data.frame()
      rownames(dat) <- dat$cell
      colData(train_sce)$cell_type <- factor(dat[colnames(train_sce), ]$cluster)
      simu_sce <- scdesign3(sce = train_sce, 
                            celltype = 'cell_type', 
                            pseudotime = NULL, 
                            spatial = NULL, 
                            other_covariates = NULL, 
                            corr_formula = "1", 
                            mu_formula = "cell_type", 
                            sigma_formula = "cell_type", 
                            n_cores = 2, 
                            copula = "gaussian", 
                            assay_use = "counts", 
                            family_use = "nb")
      
      simu_sce
    })
```


## Visualization
After the simulations, we first extract out scDesign3's BIC, which is an unsupervised metric for evaluating the goodness-of-fit of the clustering labels. 
```{r, message=TRUE}
bic_list <- lapply(scDesign3_result, function(x){return(x$model_bic)})
kmeans_ari <- sapply(kmeans_res, function(x){ARI(x$cluster, x$trueclass)})
bic_df <- data.frame(matrix(unlist(bic_list), nrow = length(bic_list), byrow = TRUE))
colnames(bic_df) <- names(bic_list[[1]])
rownames(bic_df) <- paste0("k = ", 2:10)
bic_df
```

Since we also have ground truth cell type labels from `DuoClustering2018`, we also calculate the Adjusted Rand Index (ARI) using each set of clustering labels from k-means clustering and the ground truth cell type labels. The ARI is a supervised metric to evaluate the clustering qualities. The figure below demonstrates that scDesign3's BIC agrees with ARI. 
```{r, message=FALSE}

metric <- tibble(ari = kmeans_ari, bic = bic_df$bic.marginal, Method = paste0("k = ", 2:10))
p_cluster_metric <- metric %>% ggplot(aes(x =ari, y = bic,label = Method)) + geom_point() + theme_bw() + theme(aspect.ratio = 1,
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank()) + ggpubr::stat_cor(method = "spearman", cor.coef.name = "rho", label.x.npc = "left", label.y.npc = 0.5) + ylab("scDesign3 BIC") + xlab("ARI")
p_cluster_metric
```



## Session information
```{r}
sessionInfo()
```
